Racets: Embedding Policy{-}Agnostic Programming in Racket

### 2018-07-20 paper/paper.tex
partial work on 4

\usepackage{amsmath}
\usepackage{listings}
\usepackage{racketlst}
\usepackage{xspace}
\lstset{basicstyle=\ttfamily,language=Racket,xleftmargin=.1\textwidth}
\usepackage{letltxmacro}
\newcommand*{\SavedLstInline}{}
\LetLtxMacro\SavedLstInline\lstinline
\DeclareRobustCommand*{\lstinline}{%
  \ifmmode
    \let\SavedBGroup\bgroup
    \def\bgroup{%
      \let\bgroup\SavedBGroup
      \hbox\bgroup
    }%
  \fi
  \SavedLstInline
}

@@

\section{A Formal Semantics for Faceted Execution}

@@

          \inferrule[\textsc{Apply}]{e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', v_1 \and
                                     e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v_2 \\\\
                                     \fcol{(v_1 ~ v_2), \rho, \sigma'' \Downarrow_{pc}^A \sigma''', v'}}

@@

        \\\\

@@

                    {e, \rho, \sigma \Downarrow_{pc}^E \sigma', v \and

@@

                    {(\code{box} ~ e), \rho, \sigma \Downarrow_{pc}^E \sigma'', \alpha}

@@

                    {e, \rho, \sigma \Downarrow_{pc}^E \sigma', v \\\\

@@

                    {(\code{unbox} ~e), \rho, \sigma \Downarrow_{pc}^E \sigma', v'}

@@

                    {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', v_1 \and
                     e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v_2 \\\\

@@

            \{+\ell, -\ell\} \cap pc = \varnothing \\\\

@@

          \inferrule[\textsc{Fac-Create-Pos}]

@@

           e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v}

@@

          \inferrule[\textsc{Fac-Create-Neg}]

@@

           e_3, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v}

@@

      \end{array}
    \end{displaymath}
    \\
    \hfill \textit{(Label Creation / Observation)}
    \\
    \[
    \begin{array}{cc}
      {\color{\colorFAC}
        \inferrule[\textsc{Let-Label}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \langle \lambda x. ~ e, \rho' \rangle \\\\
           \alpha \not\in dom(\sigma') \and \sigma'' = \sigma' [\alpha \mapsto \langle \lambda x. ~ e, \rho' \rangle] \\\\
           e_2, \rho [ \ell \mapsto \alpha ], \sigma'' \Downarrow_{pc}^E \sigma''', v}
          {(\code{let-label} ~ \ell ~ e_1 ~ e_2), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{Obs}]
                    {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell
                      \and e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v  \\\\
                      \langle (\lambda x.~e), \rho' \rangle = \sigma''(\ell) \\\\
                      e, \rho' [x \mapsto v], \sigma'' \Downarrow_{pc}^E \sigma''', v^{\pm} \\\\
                      e_3, \rho, \sigma''' \Downarrow_{pc}^E \sigma'''', v' \\\\
                      v'' = \fcol{\var{obs}(\ell, v', v^{\pm})}}
                    {(\code{obs} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'''', v''}}
    \end{array}
    \]
    \\
    \hfill \textit{(Possibly-Faceted Application)} \boxed{\fcol{(v_1 ~ v_2), \rho, \sigma \Downarrow_{pc}^A \sigma, v}}
    \\
    \begingroup
    \color{\colorMATH}
    \begin{displaymath}
      \begin{array}{c}
        \begin{array}{ccc}
          {\color{\colorMATH}
          \inferrule[\textsc{App-$\star$}]
          { }
          {(\star ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma, \star}}
          & 
          {\color{\colorMATH}
          \inferrule[\textsc{App-Base}]
          {e, \rho'[x \mapsto v], \sigma \Downarrow_{pc}^E \sigma', v' }
          {(\langle \lambda x.~e, \rho' \rangle ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Split}]
          {\{+\ell, -\ell\} \cap pc = \varnothing \and
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc \cup \{+\ell\}}^A \sigma', {v^+}' \\\\
            (v^- ~ v), \rho, \sigma' \Downarrow_{pc \cup \{-\ell\}}^A \sigma'', {v^-}' \and
            v' = \formfacet{\{+ \ell\}}{{v^+}'}{{v^-}'}}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma'', v'}}
        \end{array}
        \\
        \begin{array}{cc}
        \\\\
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Pos}]
          {+\ell \in pc \\\\
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Neg}]
          {-\ell \in pc \\\\
            (v^- ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
        \end{array}
      \end{array}
    \end{displaymath}
    \endgroup
    \label{fig:semantics}
\end{figure*}

\begin{figure*}
  \hfill \textit{(Metafunctions)}
      \\
      \begin{displaymath}
        \begin{array}{rclrr}
          {read(\sigma, \alpha, pc)} & {=} & {\sigma(\alpha)} & & \\
          {read(\sigma, \facet{\ell}{v_1}{v_2}), pc} & = & {read(\sigma, v_1, pc)} & \textit{if} ~ {+\ell} \in pc & \\
          {read(\sigma, \facet{\ell}{v_1}{v_2}), pc} & = & {read(\sigma, v_2, pc)} & \textit{if} ~ {-\ell} \in pc & \\
          {read(\sigma, \facet{\ell}{v_1}{v_2}), pc} & = & {\formfacet{\ell}{read(\sigma, v_1, pc)}{read(\sigma, v_2, pc)}} & \textit{otherwise} &
        \end{array}
      \end{displaymath}
      \\
      \begin{displaymath}
        \begin{array}{rclrr}
          {write(\sigma, \alpha, pc, v)} & {=} & {\sigma[\alpha ~ := \formfacet{pc}{v}{\sigma(\alpha)}]} & & \\
          {write(\sigma, \facet{\ell}{v_1}{v_2}), pc, v} & = & {\sigma''} & \textit{where} ~ {\sigma' = write(\sigma, v_1, pc \cup \{+\ell\}, v)} \\
          & & & \textit{and} ~ {\sigma'' = write(\sigma', v_2, pc \cup \{-\ell\}, v)}
        \end{array}
      \end{displaymath}
      \\
      \begin{displaymath}
        \begin{array}{rclrr}
          {obs(\ell, b, bv)} & {=} & {bv} & & \\
          {obs(\ell, true,} \facet{\ell}{v_1}{v_2}) & = & {v_1} & & \\
          {obs(\ell, false,} \facet{\ell}{v_1}{v_2}) & = & {v_2} & & \\
          {obs(\ell, b,} \facet{\ell'}{v_1}{v_2}) & = \facet{\ell'}{obs(\ell,b,v_1)}{obs(\ell,b,v_2)} & \textit{where} & {\ell} \neq {\ell'}
        \end{array}
      \end{displaymath}
\end{figure*}

@@

\[
\begin{array}{ccc}
  {\color{brown}
    \inferrule[\textsc{Fac-Create-Left}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{+ \ell \in pc} \\\\
                e_2, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  & 
  {\color{gray}
    \inferrule[\textsc{Fac-Create-Right}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{- \ell \in pc} \\\\
                e_3, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  &
  {\color{purple}
    \inferrule[\textsc{Fac-Create-Split}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{\{+\ell, -\ell\} \cup pc = \varnothing} \\\\
                e_2, \rho, \sigma' \Downarrow_{pc \cup \{+l\}}^E \sigma'', v_1 \\\\
                e_3, \rho, \sigma'' \Downarrow_{pc \cup \{-l\}}^E \sigma''', v_2 \and
                v = \formfacet{pc}{v_1}{v_2}}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}

@@

Translating these rules to Racket involves observing that each of
these rules will apply under one of three disjoint circumstances (each
of them boxed in the above rules): $+\ell \in pc$, $-\ell \in pc$, or
else $\{+\ell,-\ell\} \cap pc = \varnothing$. This leads to a
relatively straightforward implementation that inspects the current
program counter to decide which rule to implement:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(define-syntax-rule (fac l e1 e2)
  (cond
|{\color{brown}|    [(set-member? (current-pc) (pos (facet-labelname l))) e1] |}|
|{\color{gray}|    [(set-member? (current-pc) (neg (facet-labelname l))) e2]  |}|

|{\color{purple}|    [else
     (let ([left
            (parameterize ([current-pc (set-add (current-pc)
                                                (pos (facet-labelname l)))]))]
           [right
            (parameterize ([current-pc (set-add (current-pc)
                                                (neg (facet-labelname l)))]))]
           (mkfacet (labelpair-name l) v1 v2)))])) |}|
\end{lstlisting}

As our definition is a transliteration of that in Figure 6 of Austin
et. al's paper~\cite{Austin:2012}, we omit the implementation here.
The result is a function, \code{mkfacet}, which (given a label and
both positive and negative branches) forms a facet:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(define (mkfacet label v1 v2)
  (canonicalize (set-add (current-pc) (pos name)) v1 v2))
\end{lstlisting}

>>>>

\usepackage{amsmath}
\usepackage{listings}
\usepackage{racketlst}
\usepackage{xspace}
\lstset{language=Racket,xleftmargin=.1\textwidth,escapeinside={<@}{@>}}
\usepackage{letltxmacro}
\newcommand*{\SavedLstInline}{}
\LetLtxMacro\SavedLstInline\lstinline
\DeclareRobustCommand*{\lstinline}{%
  \ifmmode
    \let\SavedBGroup\bgroup
    \def\bgroup{%
      \let\bgroup\SavedBGroup
      \hbox\bgroup
    }%
  \fi
  \SavedLstInline
}

@@

\section{A Formal Semantics for Faceted Execution}
\label{sec:semantics}

@@

          \inferrule[\textsc{Apply}]{e_1, \rho, \sigma \Downarrow_{pc}^E v_1, \sigma' \and
                                     e_2, \rho, \sigma' \Downarrow_{pc}^E v_2, \sigma'' \\\\
                                     \fcol{(v_1 v_2), \rho, \sigma'' \Downarrow_{pc}^A \sigma''', v'}}

@@

@@

                    {e, \rho, \sigma \Downarrow_{pc}^E v, \sigma' \and

@@

                    {(\code{box} ~ e), \rho, \sigma \Downarrow_{pc}^E v, \sigma''}

@@

                    {e, \rho, \sigma \Downarrow_{pc}^E v, \sigma' \\\\

@@

                    {(\code{unbox} ~e), \rho, \sigma \Downarrow_{pc}^E v', \sigma'}

@@

                    {e_1, \rho, \sigma \Downarrow_{pc}^E v_1, \sigma' \and
                     e_2, \rho, \sigma' \Downarrow_{pc}^E v_2, \sigma'' \\\\

@@

            \{+\ell, -\ell\} \cup pc = \varnothing \\\\

@@

          \inferrule[\textsc{Fac-Create-Left}]

@@

           e_2, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}

@@

          \inferrule[\textsc{Fac-Create-Right}]

@@

           e_3, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}

@@

      \end{array}
    \end{displaymath}
    \\
    \hfill \textit{(Label Creation / Observation)}
    \\
    \[
    \begin{array}{cc}
      {\color{\colorFAC}
        \inferrule[\textsc{Let-Label}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \langle \lambda x. ~ e, \rho' \rangle \\\\
           \alpha \not\in dom(\sigma') \and \sigma'' = \sigma' [\alpha \mapsto \langle \lambda x. ~ e, \rho' \rangle] \\\\
           e_2, \rho [ \ell \mapsto \alpha ], \sigma'' \Downarrow_{pc}^E \sigma''', v}
          {(\code{let-label} ~ \ell ~ e_1 ~ e_2), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{Obs}]
                    {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell
                      \and e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v  \\\\
                      \langle (\lambda x.~e), \rho' \rangle = \sigma''(\ell) \\\\
                      e, \rho' [x \mapsto v], \sigma'' \Downarrow_{pc}^E \sigma''', v^{\pm} \\\\
                      e_3, \rho, \sigma''' \Downarrow_{pc}^E \sigma'''', v' \\\\
                      v'' = \fcol{\var{obs}(\ell, pc, v', v^{\pm})}}
                    {(\code{obs} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'''', v''}}
    \end{array}
    \]
    \\
    \hfill \textit{(Possibly-Faceted Application)} \boxed{\fcol{(v ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma, v}}
    \\
    \begingroup
    \color{\colorMATH}
    \begin{displaymath}
      \begin{array}{c}
        \begin{array}{ccc}
          {\color{\colorMATH}
          \inferrule[\textsc{App-$\star$}]
          { }
          {(\star ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma, \star}}
          & 
          {\color{\colorMATH}
          \inferrule[\textsc{App-Base}]
          {e, \rho'[x \mapsto v], \sigma \Downarrow_{pc}^E \sigma', v' }
          {(\langle \lambda x.~e, \rho' \rangle ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Split}]
          {\{+\ell, -\ell\} \cap pc = \varnothing \and
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc \cup \{+\ell\}}^A \sigma', {v^+}' \\\\
            (v^- ~ v), \rho, \sigma' \Downarrow_{pc \cup \{-\ell\}}^A \sigma'', {v^-}' \and
            v' = \formfacet{\{+ \ell\}}{{v^+}'}{{v^-}'}}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma'', v'}}
        \end{array}
        \\
        \begin{array}{cc}
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Pos}]
          {+\ell \in pc \\\\
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Neg}]
          {-\ell \in pc \\\\
            (v^- ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
        \end{array}
      \end{array}
    \end{displaymath}
              %%\inferrule[]
    \endgroup
    \label{fig:semantics}
\end{figure*}

@@

\[\small
\begin{array}{ccc}
  {\color{brown}
    \inferrule[\textsc{Fac-Create-Pos}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{+ \ell \in pc} \\\\
                e_2, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  & 
  {\color{gray}
    \inferrule[\textsc{Fac-Create-Neg}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{- \ell \in pc} \\\\
                e_3, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  &
  {\color{purple}
    \inferrule[\textsc{Fac-Create-Split}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{\{+\ell, -\ell\} \cup pc = \varnothing} \\\\
                e_2, \rho, \sigma' \Downarrow_{pc \cup \{+l\}}^E \sigma'', v_1 \\\\
                e_3, \rho, \sigma'' \Downarrow_{pc \cup \{-l\}}^E \sigma''', v_2 \and
                v = \formfacet{pc}{v_1}{v_2}}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}

@@

Translating these rules to Racket involves observing that each one
will apply under one of three disjoint circumstances (each of them
boxed in the above rules): \bmth{+\ell \in pc}, \bmth{-\ell \in pc},
or else \bmth{\{+\ell,-\ell\} \cap pc = \varnothing}. This is a common
idiom in our faceted semantics, as we often want to select the
appropriate branch of a facet if its label already exists in
\bmth{pc}.

At first glance, it may not be obvious that we even \emph{need} a maco
for facet creation. But according to our semantics, the following
snippet should produce \code{#t} if \bmth{+\ell \in pc}:

\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=other,numbers=none]
(fac l #t (error ''this shouldn't get evaluated if <@\bmth{+\ell \in pc}@>''))
\end{lstlisting}

If we were to implement \code{fac} as a function, it would force
evaluation of the negative branch, inconsistent with the semantics of
\textsc{Fac-Create-Pos}. We can implement each of these conditions as
a Racket macro by considering whether \bmth{\ell \in pc}, as shown in
Figure~\ref{fig:fac-create}. Each color in the listing corresponds to
the analogous semantic rule. The implementation of
\textsc{Fac-Create-Pos} and \textsc{Fac-Create-Neg} is relatively
straightforward, but \textsc{Fac-Create-Split} extends \bmth{pc} for
each branch and subsequently forms a facet. This function implements
canonicalizing facet construction, and (as the implementation is a
transliteration of that in Figure 6 of Austin et
al.~\cite{Austin:2012}) we omit its definition here.

\begin{figure}
{\small
\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=racets]
(define-syntax-rule (fac l e1 e2)
  (cond
\end{lstlisting}
\vspace{-.7\baselineskip}
{\color{brown}
\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=racets,identifierstyle=\color{brown}]
    [(set-member? (current-pc) (pos (facet-labelname l))) e1]
\end{lstlisting}
}
\vspace{-.7\baselineskip}
{\color{gray}
\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=racets,identifierstyle=\color{gray}]
    [(set-member? (current-pc) (neg (facet-labelname l))) e2]
\end{lstlisting}
}
\vspace{-.7\baselineskip}
{\small\color{purple}
\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=racets,identifierstyle=\color{purple},keywordstyle=\color{purple}]
    [else
     (let ([left (parameterize
                     ([current-pc (set-add (current-pc)
                                           (pos (facet-labelname l)))]))]
           [right (parameterize
                      ([current-pc (set-add (current-pc)
                                            (neg (facet-labelname l)))]))]
           (mkfacet (labelpair-name l) v1 v2)))]))
\end{lstlisting}}
\caption{Facet creation as a macro}
\label{fig:fac-create}
\end{figure}

\paragraph*{Label Creation}

As we chose a representation of labels as pairs of symbols (the
label's name) and closures (the predicate corresponding to the label),
label creation is relatively straightforward from the semantics:

\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=racets]
(define-syntax-rule (let-label l (lambda xs e) body)
  (let ([l (labelpair (gensym 'lab)
                      (lambda xs e))])
    body))
\end{lstlisting}

\paragraph*{Faceted References}

Our faceted semantics includes explicit \code{box} and \code{unbox}
forms. This differs from Racket's semantics, where any variable may be
treated as a box due to assignment conversion. We have two main
options:

\begin{itemize}
\item Introduce an explicit \code{unbox} form in Racets, trusting the
  programmer to explicitly use our implementation of \code{unbox} on
  potentially-faceted objects.

\item Walk over Racket code (after macro-expansion via
  \code{local-expand}) transforming variable references to use
  explicit \code{unbox} forms from Racets.
\end{itemize}

For our prototype of Racets, we have chosen to implement the
first. This leads to a relatively simple implementation, but
essentially trusts the programmer to use Racets' \code{unbox} forms
when necessary.

\begin{figure}
(define-syntax (ref-set! stx)
  (syntax-case stx ()
    [(_ var e)
     #`(let ([value e])
         (let setf ([var var]
                    [pc (current-pc)])
           (if (box? var)
               ; if var is a facet value
               (set-box! var (construct-facet-optimized (set->list (current-pc)) value (unbox var)))
               ; else
               (mkfacet
                (facet-labelname (unbox var))
                (setf
                 (facet-left (unbox var))
                 (set-add pc (pos (facet-labelname var))))
                (setf
                 (facet-right (unbox var))
                 (set-add pc (neg (facet-labelname var))))))))]))
\caption{Racets' implementation of \code{set!}}
\label{fig:setbang}
\end{figure}

Racets defines its \code{box} macro, along with \code{unbox} and
\code{set!}. We include the definition of \code{set!} in
Figure~\ref{fig:setbang}, which inlines the definition of the
\textit{write} metafunction from Section~\ref{sec:semantics} to
consider the case under which a facet is used when an address is
expected.
